---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
```

# covsel

The *covsel* R package is a ready-to-use, automated, covariate selection tool for species distribution modelling. It implements and streamlines the two steps of our novel “embedded” covariate selection procedure that combines (Step A) a collinearity-filtering algorithm and (Step B) three model-specific embedded regularization techniques, including generalized linear model with elastic net regularization, generalized additive model with null-space penalization, and guided regularized random forest. More details will come in the companion paper by [hidden for peer-review] (in prep).

### Installation

The *covsel* package requires a standard installation of R (version≥4.0.0). You can install the development version of *covsel* from GitHub:

```{r Install covsel, warning = FALSE, message = FALSE}
if(!"covsel" %in% installed.packages()) devtools::install_github("N-SDM/covsel")
```

## Package functionalities

The current version of the *covsel* package (ver. 1.0) includes a set of three functions. See function help files for additional details on input data, arguments, and examples.

| Function name           | Function description                                                                                                        |
|-------------------------|-----------------------------------------------------------------------------------------------------------------------------|
| `covsel.filteralgo()`   | Collinearity filtering (Step A)                                                                                                     |
| `covsel.embed()`        | Model-specific embedding (Step B)                                                                                                   |
| `covsel.filter()`       |Wrapper function applying the collinearity filtering algorithm at each target level(s) (e.g. i: variable level; ii: category level; iii: all remainders)|

# Example

### Introduction

This example illustrates the functionalities of the *covsel* package. The goal is to model the habitat suitability of the alpine marmot (*Marmota marmota*) in Switzerland, starting with a suite of 75 candidate covariates derived from 8 main categories (bioclimatic, land use and cover, edaphic, hydrologic, human population, transportation, vegetation, and topographic). Our aim is to reduce the dimensionality of the covariate set and select the top 12 covariates that will be used in the final model.

```{r setup, warning = FALSE, eval = TRUE}
library(covsel)
```

### Load data

The *data_covsel* dataset attached to the *covsel* package contains a `list` of three objects: (i) `data_covsel$pa` a numeric vector of presences '1' and absences '0', (ii) `data_covsel$env_vars` a data frame containing covariate data extracted at `pa` locations, and (iii) `data_covsel$catvar` a two columns look-up data frame `data_covsel$catvar$variable` and `data_covsel$catvar$category`, the variable-level names and category-level names of each covariate, respectively. Information on variable- and category-level covariate names will be useful for applying the collinearity filtering algorithm in a stratified way (e.g.: variable level first, then category level, then all remainders).

```{r explore sample data, warning = FALSE, eval = TRUE}
table(data_covsel$pa) # 3,609 presences and 10,000 background absences
dim(data_covsel$env_vars) # 75 candidate environmental covariates extracted at pa locations
```


### Covariate selection

The selection procedure is made of two main steps: (Step A) “Collinearity filtering”, and (Step B) “Model-specific embedding”.

#### Step A: Collinearity filtering

In Step A, we reduce the dimensionality of the candidate covariate set by eliminating the less informative covariates among collinear pairs. This is done by iteratively reducing a correlation matrix in which the covariates are ordered based on univariate GLM p-values. Collinear covariate pairs are identified using the Pearson correlation coefficient |r| threshold corcut, with corcut > 0.70 as default value. For maximizing the diversity of selected covariates, the filtering step can be sequentially applied at three levels: (i) the variable (e.g., selecting the best covariate for the “proportion of forest” variable calculated in 100-m, 500-m, or 1-km radii), (ii) the category (e.g.: within the “bioclimatic”, “edaphic”, or “hydrologic” categories), and (iii) using all remainders.

In this example, will run the collinearity filtering algorithm (i) first directly on the whole set of candidate covariates, (ii) then in a sequential way (variable level first, then category level, then all remainders).

##### directly on the whole set

For running the collinearity filtering algorithm on the whole candidate covariate set, we directly use the `covsel.filteralgo` function. Here we are using the default `corcut` value |r| < 0.70 for identifying collinear pairs. It is possible to assign weights to each element in the `pa` vector and the argument `force` can be used to specify a character vector indicating the name(s) of the covariate(s) to be forced in the final set. See help(`covsel.filteralgo`) for details.

```{r collinearity filtering whole set, warning = FALSE, eval = TRUE}
covdata<-data_covsel$env_vars
pa<-data_covsel$pa
dim(covdata) # 75 candidates before collinearity filtering
covdata_filter<-covsel.filteralgo(covdata=covdata,
                                  pa=pa,
                                  corcut=0.7) # default value
dim(covdata_filter) # much less after
```

##### in a sequential way

For running the collinearity filtering algorithm in a sequential way (e.g.: variable level first, then category level, then all remainders), we will use the wrapper function `covsel.filter`. In addition to the arguments described for `covsel.filteralgo`, this function requires information on the variable-level or/and category-level covariate names. These can be provided as character vectors to the arguments `variables` and `categories`, respectively.

```{r collinearity filtering stratified, warning = FALSE, eval = TRUE}
covdata<-data_covsel$env_vars
pa<-data_covsel$pa
dim(covdata) # 75 candidates before collinearity filtering
covdata_filter<-covsel.filter(covdata=covdata,
                              pa=pa,
                              corcut=0.7, # default value
                              variables=data_covsel$catvar$variable,
                              categories=data_covsel$catvar$category)
dim(covdata_filter) # much less after
```


#### Step B: Model-specific embedding

Covariates selected after step A are used to fit models with embedded selection procedures using the `covsel.embed` function. Available algorithms in the current version of the *covsel* package (ver. 1.0) are: GLM with elastic-net regularization, GAM with null-space penalization, and guided regularized RF. They can be used together (default), or individually by tuning the `algorithms` argument. For each algorithm, the n covariates retained after regularization are ranked from 1 (“best”) to n (“worst”). The algorithm-specific ranking is done based on the absolute value of the regularized regression coefficients for GLM, the chi-square statistic for GAM, and the Mean Decrease Gini index for RF, all to be maximized. The final overall ranking is obtained by ordering the sum of the three ranks for each covariate, starting with the covariates that were commonly selected by all the algorithms, and then adding the remaining. The top ncov covariates are selected as the final modelling set, with `ncov` and `maxncov` being user-specifiable arguments with default values round(log2(number of occurrences)) and 12, respectively. See help(`covsel.embed`) for details on other optional arguments (i.e. `weights`, `force`, `nthreads`, etc.).

```{r embedding, warning = FALSE, eval = TRUE}
covdata<-data_covfilter
pa<-data_covsel$pa
dim(covdata) # 37 candidates before embedding
covdata_embed<-covsel.embed(covdata=covdata,
                            pa=pa,
                            algorithms=c('glm','gam','rf'), # default value
                            ncov=ceiling(log2(length(which(pa==1)))), # default value
                            maxncov=12, # default value
                            nthreads=detectCores()/2)  # default value
```

Here we are! Below is the list of the top 12 covariates retained for modelling the habitat suitability of the alpine marmot (*Marmota marmota*) in Switzerland.

```{r results, eval = TRUE}
dim(covdata_embed$covdata) # data.frame with the top 12 retained for the final modelling set
print(covdata_embed$ranks_2) # ranking table
```

# Contributing

If you have a suggestion that would make *covsel* better, please fork the repository and create a pull request. You can also simply open an issue with the tag "enhancement". Thanks!

# Citation

To cite *covsel* or acknowledge its use, cite the package as follows, substituting the version of *covsel* that you used for “version 1.0”:

[info hidden for peer review] et al. 2022. Too many candidates: embedded covariate selection procedure for species distribution modelling with the covsel R package. – XXX XXX: XXX (ver. 1.0).

# Contact

[info hidden for peer review]

Project Link: https://github.com/N-SDM/covsel

# Acknowledgments

*covsel* development has been conducted within the [info hidden for peer review] lab

We gratefully acknowledge financial support through the Action Plan of the Swiss Biodiversity Strategy by the Federal Office for the Environment (FOEN) for financing the Valpar.ch and SwissCatchment projects.

The Swiss Species Information Center InfoSpecies (www.infospecies.ch) supplied Swiss-level species occurrence data and expertise on species’ ecology, and we acknowledge their support regarding the database.

This research was enabled in part by the support provided by the Scientific Computing and Research Unit of Lausanne University (https://www.unil.ch/ci/dcsr).
